\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-08-07 analysis/paper/paper.Rmd

<!--Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison. The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years. Therefore, we will examine three Cox regression scenarios: sps only, aps only and all covariates excluding sps and aps. Next, we will compare each modeling using the Brier score [@graf1999ass]. The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context. For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability. Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions. Figure \ref{fig:brier1} examines the brier scores as they change over time. As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.-->

@@

For our penalized logistic regression model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}. In order to use this interface, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space. We used lasso penalization by setting \code{alpha = 1}. Moreover, we use the argument \code{penalty.factor} to force the time variable into the final model.

@@

abCB <- data.frame(Time = abCbFull[, 1],
                   absRisk = rowMeans(abCbFull[, -c(1)]))
abCB$Model <- "Penalized logistic"

@@

Our first comparison plot examines the covariates that were selected between all three hazard functions. Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have larger estimated coefficients. This is expected, as there are no restrictions on the size of coefficients. The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).

@@

estimatescb <- data.table::setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
                     keep.rownames = TRUE)
# estimatescb$std.error<-0
estimatescb$Model <- "Pen. logistic"
colnames(estimatescb) <- c("term", "estimate", "model")

@@

These values do not inform us of model performance. To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time. The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). There is separation between the curves, but it is minimal. To better demonstrate the differences in performance between the models, we compare Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models. The differences, however, are minimal. All three outperform the Kaplan-Meier model.

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
  geom_line(aes(y = absRisk, colour = Model)) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=9)
```

@@

```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. All follow-up times seen in the new data were used. (B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. Lower is better. Penalized logistic regression produced the lowest results over follow-up time.", warning = FALSE}
# Compute Brier score for penalized models
brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized Logistic" = cbFull),
                           data = cbind(subset(test, select = c(d.time, death)),
                                        as.data.frame(newx)), 
                           formula = Hist(d.time, death != 0) ~ 1, summary = NULL, 
                           se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
#brierFinalResults$models<-as.character(brierFinalResults$models)
# Combine Brier scores between unpenalized and penalized models before plotting

>>>>

<!--Before generating an absolute risk curve using regularized estimation, we will explore three options as a baseline comparison. The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years. Therefore, we will examine three Cox regression scenarios: sps only, aps only and all covariates excluding sps and aps. Next, we will compare each modeling using the Brier score [@graf1999ass]. The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context. For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability. Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions. Figure \ref{fig:brier1} examines the brier scores as they change over time. As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized estimation and the unadjusted Kaplan-Meier curve.-->

@@

For our penalized model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}. In order to use this interface, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space. We used lasso penalization by setting \code{alpha = 1}. Moreover, we use the argument \code{penalty.factor} to force the time variable into the final model.

@@

abCB <- data.frame(Time = abCbFull[, 1],
                   absRisk = rowMeans(abCbFull[, -c(1)]))
abCB$Model <- "Penalized estimation"

@@

Our first comparison plot examines the covariates that were selected between all three hazard functions. Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have larger estimated coefficients. This is expected, as there are no restrictions on the size of coefficients. The smallest coefficient estimates are found in penalized case-base model, aside from cancer state (cano).

@@

estimatescb <- data.table::setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
                     keep.rownames = TRUE)
# estimatescb$std.error<-0
estimatescb$Model <- "Pen. CB"
colnames(estimatescb) <- c("term", "estimate", "model")

@@

These values do not inform us of model performance. To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time. The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). There is separation between the curves, but it is minimal. To better demonstrate the differences in performance between the models, we compare Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized case-base model compared to both Cox regression models. The differences, however, are minimal. All three outperform the Kaplan-Meier estimator.

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
  geom_line(aes(y = absRisk, colour = Model)) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized CB")) + 
  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=9)
```

@@

```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier. (A)  Probabilities of death as it changes over follow-up time. (B) Brier scores as they change over time, where a lower score corresponds to better performace. Penalized case-base estimation produced the lowest results over follow-up time.", warning = FALSE}
# Compute Brier score for penalized models
brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized CB" = cbFull),
                           data = cbind(subset(test, select = c(d.time, death)),
                                        as.data.frame(newx)), 
                           formula = Hist(d.time, death != 0) ~ 1, summary = NULL, 
                           se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
#brierFinalResults$models<-as.character(brierFinalResults$models)
# Combine Brier scores between unpenalized and penalized models before plotting

### 2020-08-07 analysis/paper/paper.tex

These values do not inform us of model performance. To do so, we take
our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
curves, but it is minimal. To better demonstrate the differences in
performance between the models, we compare Brier scores over time. In
Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
penalized logistic regression model compared to both Cox regression
models. The differences, however, are minimal. All three outperform the
Kaplan-Meier model.

@@

\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. All follow-up times seen in the new data were used. (B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. Lower is better. Penalized logistic regression produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

@@

This report was generated on 2020-08-06 17:18:33 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.6
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.5.6               dotwhisker_0.5.0         
#>  [3] timereg_1.9.4             glue_1.4.1               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.0              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.0.2               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.8          
#> [15] visreg_2.6.1              tibble_3.0.3             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.1-12          
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.1.0          glmnet_3.0-2             
#> [25] Matrix_1.2-18             data.table_1.12.8        
#> [27] cowplot_1.0.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
#> [73] withr_2.2.0         pillar_1.4.6        haven_2.2.0        
#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
#> [79] mstate_0.2.12       modelr_0.1.6        crayon_1.3.4       
#> [82] rmarkdown_2.1       jpeg_0.1-8.1        grid_4.0.2         
#> [85] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
#> [88] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
#> [91] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

>>>>

These values do not inform us of model performance. To do so, we take
our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
curves, but it is minimal. To better demonstrate the differences in
performance between the models, we compare Brier scores over time. In
Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
penalized case-base model compared to both Cox regression models. The
differences, however, are minimal. All three outperform the Kaplan-Meier
estimator.

@@

\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier. (A)  Probabilities of death as it changes over follow-up time. (B) Brier scores as they change over time, where a lower score corresponds to better performace. Penalized case-base estimation produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

@@

This report was generated on 2020-08-06 17:31:23 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.6
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.5.6               dotwhisker_0.5.0         
#>  [3] timereg_1.9.4             glue_1.4.1               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.0              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.0.2               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.8          
#> [15] visreg_2.6.1              tibble_3.0.3             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.1-12          
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.1.0          glmnet_3.0-2             
#> [25] Matrix_1.2-18             data.table_1.12.8        
#> [27] cowplot_1.0.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
#> [73] pillar_1.4.6        haven_2.2.0         foreign_0.8-80     
#> [76] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
#> [82] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
#> [88] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
#> [91] viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}
